% !TEX root = ../main.tex

\section{Gaps discovered during formalization}\label{app:gaps}

Formalizing the AstroNet axioms and depth filtration in Lean~4 with Mathlib uncovered four issues, all corrected in the Theory module:
\begin{enumerate}
  \item[(a)] \emph{Atom self-loop in cycle definition.} Atoms with $\mathrm{ref} = (\mathrm{id}(e))$ trivially satisfy a na\"ive cycle predicate. The formalization required a structured path definition (\texttt{AstroPath}) with a distinct-node condition.
  \item[(b)] \emph{Missing no-self-reference axiom.} AstroNerves with $\omega(e) > 0$ and $\mathrm{id}(e) \in \mathrm{ref}(e)$ break the cycle characterization. Added as the fifth axiom (Definition~\ref{def:well-formed}, condition~4).
  \item[(c)] \emph{Redundant non-emptiness axiom.} $|\mathrm{ref}(e)| \geq 1$ follows from self-reference and width. Removed from the axiom list and derived as a lemma (\texttt{AstroNet.nonEmpty}).
  \item[(d)] \emph{Stabilization theorem strengthening.} The original biconditional (``unstaged if and only if on a cycle'') is false as stated; corrected to two separate implications: cycle nerves are unstaged (\texttt{depthFilt\_not\_inCycle}), and unstaged nerves reach a cycle (\texttt{unstaged\_reaches\_cycle}).
\end{enumerate}

\section{Codebase and benchmark data}\label{app:codebase}

\begin{table}[H]
\centering\small
\begin{tabular}{lrl}
  \toprule
  Module & Lines & Contents \\
  \midrule
  GenericKernel & 397 & Parameterized CIC kernel \\
  TreeKernel & 50 & Key = String aliases \\
  NerveKernel & 563 & Hash, AstroNerve, AstroNet, ExprWithHash, Shatter \\
  Theory & 534 & AstroNet axioms, depth filtration, cycle analysis \\
  Examples & ${\sim}800$ & StoreEquivalence, Benchmark, Diagnostics, exports \\
  \bottomrule
\end{tabular}
\caption{Codebase summary.}
\label{tab:codebase}
\end{table}

\begin{table}[H]
\centering
\begin{tabular}{@{}lrrr@{}}
\toprule
Depth & GenericKernel.defEq & defEqWithHash & Speedup \\
\midrule
100 & 423\,ms & 0.32\,ms & $1{,}322\times$ \\
1{,}000 & 4{,}243\,ms & 0.37\,ms & $11{,}467\times$ \\
10{,}000 & 43{,}306\,ms & 0.30\,ms & $144{,}353\times$ \\
\bottomrule
\end{tabular}
\caption{Definitional equality benchmark (1000 iterations, two distinct allocations to defeat pointer equality).  GenericKernel.defEq is O(n) in expression depth.  defEqWithHash is O(1) via cached hash comparison.  Lean~4's production kernel achieves the same O(1) behavior via \texttt{mkData(mixHash ...)} and hash consing (\texttt{ShareCommon.shareCommon'}).}
\label{tab:benchmark}
\end{table}
